Dependent Types and Martin-L\"of Type Theory: An Introduction

### 2016-07-14 paper.tex
first commit from ubuntu

\title{Dependent Types and Martin-L\"of Type Theory: An Introduction}
\author{Carl Factora}
\date{\today}

@@

\section{Introduction}
We present an introduction to dependent types and fundamental type theory.
At its core, type theory seeks to serve as a possible foundation of mathematics,
which itself is based on a different logical foundation. This proof-relevant
form of mathematics is the basis in which types can be used to model logical
propositions and is the core idea of type theory. As such, not only does type
theory have a viable application in mathematics but also in computer science,
providing a richer world of types and expressibility.

The study of type theory goes hand in hand with the study of dependent types,
which is the reason why first take the time to do so in Section 2. After we
have derived a proper foundation for type theory, we then continue with the
foundational concepts of Martin-L\"of type theory. We recommend using a proof
assistant while working and studying with the examples we provide. For such
examples, we use Agda notation.

\section{Dependent Types}
In this section, we introduce the concept of dependent types and their role in
type theoretic systems. At their core, dependent types can be thought of simply
as an \textit{interplay of types and terms}. To understand this relationship,
it helps to think of the ways in which types and terms can \textit{depend} on
each other. Altogether, there are four different kinds of dependencies:

@@

What follows is a brief description of each form of dependency and the
corresponding \(\lambda\)-calculus that is associated with it. With this, we
introduce the (fully-dependent) system of the \textit{Calculus of Constructions}
(\(\lambda C\)). Here, the word \textit{fully} signifies that \(\lambda C\)
features all four forms of dependency mentioned above.

\subsection*{Simply-Typed (STLC)}
\subsubsection*{Role: Base System}
The STLC is not a dependently typed system. In fact, the term {\em dependent type}
is drawn in constrast with the term {\em simple type}, where the STLC derives
its namesake. A simple type is either a {\em primitive type}, which is the set
of all types ``pre-defined'' within the system, or a {\em function type} between
other simple types. Thus, simple types have the following grammar:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
t || t → t
\end{code}
\end{minipage}
\end{center}
The terms of the STLC are the terms of the \textit{untyped \(\lambda\)-calculus},
with the exception of terms that use \textit{self-application} (i.e., Y
combinator, Omega, etc.). This has the benefit of removing all \textit{infinite}
calculations from the system but also removing all recursive computations \cite{}.
Thus, the STLC is not \textit{Turing complete} and is thus the appropriate
foundation for deriving a type theoretic system. This is because the STLC lends
itself quite seamlessly as component of a larger, more expressive system. This
flexibility is what we exploit to derive \(\lambda C\). To do this, we first
extend the STLC with more expressive types through simple extentions on its
grammar for simple types.

>>>>

\title{Dependent Types and Martin-L\"of Type Theory: An Introduction}
%% \title{Dependent Types and Type Theory: An Introduction}
\author{Carl Factora}
\date{\today}

@@

\section{Introduction}
We present an introduction to dependent types and (Martin-L\"of) type theory.
At its core, type theory seeks to serve as a possible foundation of mathematics,
which itself is based on a different logical foundation. This proof-relevant
form of mathematics is the basis in which types can be used to model logical
propositions and is the core idea of type theory. As such, not only does type
theory have a viable application in mathematics but also in computer science,
providing for both a richer world of types and expressibility.

The study of type theory goes hand in hand with the study of dependent types,
which is  why we first take the time to do familiarize ourselves with them in
Section 2. We then continue with the foundational concepts of type theory of
Per Martin-L\"of \cite{}. In this section, while reading through the examples
we have provided, we recommend using a proof assistant, such as Agda. In our
examples, we refer to Agda notation.

\section{Dependent Types}
In this section, we introduce the concept of dependent types and their role in
type theoretic systems. At their core, dependent types can be thought of simply
as an \textit{interplay of types and terms}. To understand this relationship,
it helps to think of the ways in which types and terms can \textit{depend} on
one another. Altogether, this results in four dependencies:

@@

What follows is a brief description of each form of dependency and the
corresponding \(\lambda\)-calculus that is associated with it. With this, we
introduce the (fully-dependent) system of the \textit{Calculus of Constructions}
(\(\lambda C\)) \cite{}. Here, the word \textit{fully} signifies that \(\lambda C\)
features all four forms of dependency mentioned above.

\subsection*{Simply-Typed (STLC)}
\subsubsection*{Role: Base System}
The STLC is not a dependently typed system. In fact, the term {\em dependent type}
is drawn in constrast with the term {\em simple type}, where the STLC derives
its namesake. A simple type is either a {\em primitive type}, which is the set
of all types ``pre-defined'' (e.g. lists, numbers, booleans), or a
{\em function type} between other simple types. Thus, simple types have the
following grammar:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
t || t → t
\end{code}
\end{minipage}
\end{center}
The terms of the STLC are the terms of the \textit{untyped \(\lambda\)-calculus},
with the exception of terms that use \textit{self-application} (i.e., Y
combinator, Omega, etc.). An example term, annotated with its type, is generally
written as follows:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
example : ℕ → ℕ
example = λ (z : ℕ) → z
\end{code}
\end{minipage}
\end{center}
Or, equivalently,
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
example = λ z → z
\end{code}
\end{minipage}
\end{center}
whenever the type of the argument (viz. {\tt z}) is easily inferable.

This leaves the STLC rather small and limited but has the benefit of removing
all \textit{infinite} calculations from the system \cite{}. Thus, the STLC is
not \textit{Turing complete} and is thus the appropriate foundation for
deriving a type theoretic system. This is because the STLC lends itself quite
seamlessly as component of a larger, more expressive system. This flexibility
is what we exploit to derive \(\lambda C\). To do this, we first extend the STLC
with more expressive types through simple extentions on its grammar for simple
types.
